Staged Abstract Interpreters

### 2019-08-14 paper/20_prelim.tex
update

\paragraph{Primitive Operations} Next we define several primitive operations.
Two versions of @alloc@ are declared. The first one takes a store and an
identifier and produces a fresh address of non-monadic type @R[Addr]@. Since
the freshness of the address may depend on the store, which might be a
next-stage value as indicated by its type, the type of addresses is therefore
wrapped by @R[_]@. The other @alloc@ is simply the monadic version of the addresses.

@@

\paragraph{The Interpreter} Now we construct the semantics-agnostic interpreter
in monadic form, shown in Figure \ref{fig:shared_int}. The basic idea
of a generic interpretation is to traverse the abstract syntax tree
while maintaining the effects such as reader and state.  Note that the
interpreter is written in open-recursive style -- it can not refer to
itself directly, instead, @eval@ takes an additional parameter @ev@ of
type @Expr => Ans@ that refers to itself. Consequently, the method
@close@ that lifts lambda terms to closures and method @ap_clo@ that
applies functions also takes an extra @ev@, because evaluation may
happen inside of them.
To close the open-recursion, we use a fixed-point opertor @fix@.
For concrete-interpretation instantiation, it works like the Y
combinator; for abstract-interpretation instantiation, it instruments
the interpreter by memorizing its input and output, which will ensure
the termination of abstract interpretation.
Finally, a top-level wrapper @run@ is declared; the return type
@Result@ depends on what kind of monad we will be using, this is also
an abstract type.

@@

Now we can instantiate the interpreter as a standard environment-store
interpreter. Such interpreter also can be obtained by refunctionalization of
CESK machines \cite{Felleisen:1987:CAH:41625.41654, DBLP:conf/ppdp/AgerBDM03}.

\paragraph{Concrete Components}
The two types we need to concretize are @Addr@ and @Value@, afterward, the @Env@
and @Store@ is concretized automatically. To assure the freshness, we use @Int@
for the address space @Addr@. A value can be either a tagged number
\texttt{IntV}, or a closure \texttt{CloV} that contains a lambda term and an
environment. The final return value of the interpreter is a pair of values and
stores, where the @Value@ and @Store@ are obviously next-stage objects. We also
provide a standard fixed-point combinator to close the open-recursive function @ev@.

\begin{lstlisting}
  trait ConcreteComponents extends Semantics {
    type Addr = Int
    sealed trait Value
    case class IntV(i: Int) extends Value
    case class CloV(λ: Lam, e: Env) extends Value
    type Result = (R[Value], R[Store])
    def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans = e => ev(fix(ev))(e)
  }
\end{lstlisting}
%\vspace{-1em}

\paragraph{Unstaged Monads}
The environment and store can be modeled by reader effect and state effect,
i.e., a Reader monad and a State monad. We combine them using monad transformers
@ReaderT@ and @StateT@.
In other words, we instantiate the @AnsM@ monad as a stack of @ReaderT@ and @StateT@
transformers\footnote{The question mark syntax is a kind projector
  \cite{kindprojector}, thus \texttt{StateT[IdM,Store,?]} is equivalent to \newline
  \texttt{(\{type M[T]=StateT[IdM,Store,T]\})\#M}}, where the @ReaderT@ is
parameterized by type @Env@, @StateT@ is parameterized by type @Store@, and the
inner-most monad @IdM@ is just an identity monad.

\begin{lstlisting}
  trait ConcreteSemantics extends ConcreteComponents {
    type R[T] = T
    type AnsM[T] = ReaderT[StateT[IdM, Store, ?], Env, T]
    ...
  }
\end{lstlisting}

Here we sketch the basic idea of unstaged @ReaderT@ and @StateT@. Readers may
refer to \cite{DBLP:conf/popl/LiangHJ95, Chiusano:2014:FPS:2688794} for more detail 
and implementation of monad transformers.
A @ReaderT@ monad transformer encapsulates the computation @R => M[A]@, where
@R@ is the reader type, @M[_]@ is a monad type. The @ReaderT@ monad returns the
transformed value of type @M[A]@. Similarly, a @StateT@ monad encapsulates the
computation @S => M[(A, S)]@, where @S@ is the state type, @M[_]@ is a monad
type. The @StateT@ monad returns the transformed value of type @M[(A, S)]@,
where the new state is accompanied.
Note that the binding-time type @R@ is an identity type, thus the monads are
operated on unstaged data. We can also see this from the signature of @flatMap@:
the function @f@ takes an unstaged value of type @A@ and produces a monadic
value.

\begin{lstlisting}
  case class ReaderT[M[_]: Monad, R, A](run: R => M[A]) {
    def flatMap[B](f: A => ReaderT[M, R, B]): ReaderT[M, R, B] = ... }
  case class StateT[M[_]: Monad, S, A](run: S => M[(A, S)]) {
    def flatMap[B](f: A => StateT[M, S, B]): StateT[M, S, B] = ... }
\end{lstlisting}

Then, the monadic operations that manipulate the environment and store are
simply constructing the proper monads and lifting it to the top-level of our
monad stack. To modify the store, for example, we may have a @StateT@ monad that
transforms the current store $\sigma$ to @σ + αv@, which results in
@StateT[IdM, Store, Unit]@, and then lift this @StateT@ value to @ReaderT@,
i.e., @AnsM[Unit]@.
\begin{lstlisting}
  def set_store(αv: (Addr, Value)): AnsM[Unit] = liftM(StateTMonad.mod(σ => σ + αv))
\end{lstlisting}

\paragraph{Primitive Operations}
Other primitive operations can be implemented straightforwardly, for example,
@alloc@ simply takes the size of the current store and produces the successor number.
We elide most of them but describe a little bit on how do we handle functions
and applications as they will be very different when staging and abstract
interpretation is involved. The method @close@ takes a lambda term and an
environment, then create a closure value @CloV@.

@@

To apply a function, the method @ap_clo@ takes a function value and an argument
value, and extracts the lambda term and environment enclosed in the function value,
and evaluates the body expression of that lambda term under the new environment and store.

@@

With other primitive operations implemented, we can implement the top-level
@run@ method, where $\rho_0$ and $\sigma_0$ are the initial empty environment and
store.

>>>>

\paragraph{Primitive Operations} Next we define several primitive operations.
First, two versions of @alloc@ are declared. The first one takes a store and an
identifier and produces a fresh address of non-monadic type @R[Addr]@. Since
the freshness of the address may depend on the store, which might be a
next-stage value as indicated by its type, the type of addresses is therefore
wrapped by @R[_]@. The other @alloc@ is simply the monadic version of the addresses.

@@

\paragraph{The Interpreter} Now we construct the semantics-agnostic interpreter
in monadic form, shown in Figure \ref{fig:shared_int}. The basic idea
of a generic interpretation is to traverse the abstract syntax tree
while maintaining the effects such as reader and state.  Note that the
interpreter is written in open-recursive style -- it can not refer to
itself directly, instead, @eval@ takes an additional parameter @ev@ of
type @Expr => Ans@ that refers to itself. Consequently, the method
@close@ that lifts lambda terms to closures and method @ap_clo@ that
applies functions also takes an extra @ev@, because evaluation may
happen inside of them.
To close the open-recursion, we use a fixed-point opertor @fix@.
For concrete-interpretation instantiation, it works like the Y
combinator; for abstract-interpretation instantiation, it instruments
the interpreter by memoizing its input and output, which will ensure
the termination of abstract interpretation.
Finally, a top-level wrapper @run@ is declared; the return type
@Result@ depends on what kind of monad we will be using, this is also
an abstract type.

@@

As the first step in our roadmap, we instantiate the interpreter concretely in
this section. The result of such instantiation is a standard definitional
interpreter with environments and stores, it also can be obtained by
refunctionalization of CESK machines \cite{Felleisen:1987:CAH:41625.41654,
DBLP:conf/ppdp/AgerBDM03}. We first present the concrete components, i.e., the
value domain, then show the monad stack for concrete interpretation, finally
sketch the how the primitive operations are implemented.

\paragraph{Concrete Components}
The two types we need to concretize are addresses @Addr@ and values @Value@. The
type @Env@ and @Store@ is derived automatically. To assure the freshness of
address allocation, we use @Int@ and always return successor of the size of
current store. A value can be either a tagged number \texttt{IntV}, or a closure
\texttt{CloV} that contains a lambda term and an environment. The final grounded
result of the interpreter is a pair of values and stores, where the @Value@ and
@Store@ are next-stage objects, as annotated by @R@. We also define a standard
fixed-point combinator to close the open-recursive function @ev@.

\begin{lstlisting}
  trait ConcreteComponents extends Semantics {
    type Addr = Int
    sealed trait Value
    case class IntV(i: Int) extends Value
    case class CloV(λ: Lam, e: Env) extends Value
    type Result = (R[Value], R[Store])
    def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans = e => ev(fix(ev))(e)
  }
\end{lstlisting}

\paragraph{Unstaged Monads}
For the purpose of concrete interpretation, the monad needs to model read
effect and state effect, which correspond to environment and store,
respectively. Following the monad transformer approach
\cite{DBLP:conf/popl/LiangHJ95}, we use a @ReaderT@ and @StateT@ monad
transformer to compose the monad stack. In other words, the type @AnsM@ is
instantiated by layering @ReaderT@ and @StateT@ transformers\footnote{The
question mark syntax is a kind projector \cite{kindprojector}, thus
\texttt{StateT[IdM,Store,?]} is equivalent to \newline \texttt{(\{type
M[T]=StateT[IdM,Store,T]\})\#M}}, where the @ReaderT@ is parameterized by type
@Env@, @StateT@ is parameterized by type @Store@, and the inner-most monad @IdM@
is simply an identity monad.

\begin{lstlisting}
  trait ConcreteSemantics extends ConcreteComponents {
    type R[T] = T
    type AnsM[T] = ReaderT[StateT[IdM, Store, ?], Env, T] // the monad stack
    ... }
\end{lstlisting}

Here we sketch the basic idea of @ReaderT@ and @StateT@. Readers may
refer to \cite{DBLP:conf/popl/LiangHJ95, Chiusano:2014:FPS:2688794} for more detail.
A @ReaderT@ monad transformer encapsulates computation @R => M[A]@, where
@R@ is the reader type, and @M[_]@ is the inner monad type.
Given a @R@, @ReaderT@ monad produecs a transformed value of type @M[A]@.
Similarly, a @StateT@ monad encapsulates computation @S => M[(A, S)]@, where
@S@ is the state type, and @M[_]@ is the inner monad type.
Given a @S@, @StateT@ monad produces a transformed value of type @M[(A, S)]@,
where the new state (type @S@) is accompanied with the value (type @A@).
Note that for the moment, the binding-time type @R@ is an identity type, thus
these monads operate on unstaged data. We can also see this from the
signature of @flatMap@: the argument function @f@ takes an unstaged value of type @A@ and
produces a monadic value. In the following code, we elide operations other than @flatMap@.

\begin{lstlisting}
  case class ReaderT[M[_]: Monad, R, A](run: R => M[A]) {
    def flatMap[B](f: A => ReaderT[M, R, B]): ReaderT[M, R, B] =
      ReaderT(r => Monad[M].flatMap(run(r))(a => f(a).run(r))); ... }
  case class StateT[M[_]: Monad, S, A](run: S => M[(A, S)]) {
    def flatMap[B](f: A => StateT[M, S, B]): StateT[M, S, B] =
      StateT(s => Monad[M].flatMap(run(s)) { case (a, s1) => f(a).run(s1) }); ... }
\end{lstlisting}

After defining the monad stack, the operations that manipulate the environment
and store can be defined by constructing the proper monad and lifting it to the
top-level of our monad stack. To modify the store, for example, we can construct a
@StateT@ value that transforms the current store $\sigma$ to @σ + αv@, which
results in @StateT[IdM, Store, Unit]@, and then lift this @StateT@ value to the
top-level @ReaderT@ type, i.e., @AnsM[Unit]@.
\begin{lstlisting}
  def set_store(αv: (Addr, Value)): AnsM[Unit] = liftM(StateTMonad.mod(σ => σ + αv))
\end{lstlisting}

\paragraph{Primitive Operations}
Other primitive operations operates on the value domain can be implemented
straightforwardly. We elide most of them but describe a bit on how we handle
functions and applications. Because lambda terms are data but also part of the
control flow, although the way we treat them is simple now, but will be very
different when staging and abstract interpretation is involved. The method
@close@ takes a lambda term and an environment, and produces the
defunctionalized representation of closures.

@@

To apply a function, @ap_clo@ takes a function value @rator@ and an argument
@rand@, extracts the lambda term and environment enclosed in @rator@, allocates
the addresses and updates the environment and store accordingly. Finally, it
evaluates the body expression of the lambda term under the new environment and
store.

@@

The top-level @run@ method invokes the fixed-pointer operator directly,
given the expression @e@, initial environment $\rho_0$ and store $\sigma_0$.

### 2019-08-14 paper/30_stageconc.tex
update

In this section, based on the generic interpreter and concrete components we
presented in Section \ref{prelim}, we show how to obtain a staged concrete
interpreter by changing the monads and refactoring several primitive
operations. We begin by briefly introducing the Lightweight Modular Staging
framework in Scala.

>>>>

In this section, based on the generic interpreter and concrete components we
presented in Section \ref{prelim}, we show how to stage the concrete interpreter
by changing the monad type and refactoring several primitive operations. We
begin by briefly introducing the Lightweight Modular Staging framework in Scala,
and then replay the same step in the unstaged counterpart. At last, we briefly
describe code generation.
